---
title: "Finance Quantitative"
subtitle: "TP: Options sur Actions et Indices"
author: Patrick Hénaff
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    keep_tex: false
    fig_caption: yes
    latex_engine: pdflatex
    number_sections: true
geometry: margin=1in

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{eurosym}
  - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include=FALSE, echo=TRUE}
library(fOptions)
library(kableExtra)
```


# Smile Interpolation with the Vanna-Volga pricing model.

Given the implied volatility at three strikes, we will use the Vanna-Volga pricing method to interpolate the
volatility curve. Assume $r=0, b=0, T=1, \mbox{Spot} = 100$. The Black-Scholes volatility
for three strikes is given below.

```{r, echo=FALSE}
# Benchmark data: (strike, volatility)
VolData <- list(c(80, .32), c(100, .30), c(120, .315))
tmp <- as.data.frame(VolData)
row.names(tmp) <- c("Strike", "Volatility")
colnames(tmp) <- NULL
kable(tmp, booktabs=T) %>%
  kable_styling(position = "center", latex_options = "HOLD_position")
```

## Getting started

- Using the Rmetrics library (fOptions), verify that you know how to compute the price and the "greeks" of a vanilla option.
- Identify or write a robust function to compute the implied volatility, given the price of a vanilla option.

## Calculation steps


### Write a robust function to calculate the implied volatility, given a strike.

### Write functions to compute the risk indicators needed for the Vanna-Volga method. 

You may use analytical expressions or a finite difference method.


### Compute vectors of vega, vanna, volga for the three hedge instruments


### Compute the risk indicators for a call option struck at 90.

### Compute the Vanna-Volga price adjustment and the corresponding implied volatility.

### Wrap the above logic in a function in order to
Interpolate/extrapolate the vol curve from $K=70$ to $K=130$ and plot the result.


# Pricing a digital call

Recall that a digital call with strike $K$ pays one euro if $S_T \geq K$, and nothing otherwise.

The volatility curve is given in the previous question. The objective is to price a digital call, maturity $T=1$, struck at $K=105$, using the same Vanna-Volga method as in the previous question. 


## Data

```{r, echo=TRUE}
T <- 1
Spot <- 100
r <- 0
b <- 0

# Vol ATM
sigma <- .30

# strike
Strike <- 105
```

### Write a function to compute the price of a digital call analytically.

### Using the same reference instruments as in the previous question, compute the VV risk indicators of the hedge and of the digital call.


### Compute the VV price correction for the digital call


# Valorisation d'une option "chooser"

Une option "chooser" de maturité $t_2$ et strike $K$ donne le droit au détenteur de choisir si l'option est un call ou un put, à un moment $t_1$ de la vie de l'option préalablement défini. A ce moment là, l'option "chooser" vaut $\max (C(K,t_2),P(K, t_2))$, où $C(K, t_2)$ ($P(K,t_2)$) est la valeur en $t_1$ d'un call (put) de strike $K$ de maturité $t_2$.   

On se propose de valoriser cette options de deux manières: à l'aide d'un arbre trinomial d'une part, et d'autre part par une formule analytique.

## Valorisation dans un arbre trinomial

On construira un modèle trinomial par agrégation de deux pas d'un modèle binomial de type Jarrow-Rudd.

On rappelle les paramètres du modèle binomial de Jarrow-Rudd, qui est charactérisé par des probabilités de transition $p=q=1/2$:
$u = e^{\mu \Delta t + \sigma \sqrt{\Delta t}}$, 
$d = e^{\mu \Delta t - \sigma \sqrt{\Delta t}}$.

avec $\mu = (r-d) - \frac{1}{2} \sigma^2$.


### Calculer les paramètres d'un arbre trinomial constitué par l'agrégation de deux pas de temps d'un arbre binomial de Jarrow-Rudd.


### Construire un arbre de 200 pas, maturité 1 an pour le processus log-normal

$$
dS_t = r dt + \sigma dW
$$
avec les paramètres suivants:

```{r}
data = as.matrix(c(100, .20, .02), nrows=3)
rownames(data) = c("$S_0$", "$\\sigma$", "$r$")
knitr::kable(data, format="latex", booktabs=TRUE, escape=FALSE)
```


Déterminez une manière parcimonieuse de représenter cet arbre, adaptée à la résolution de la question suivante.

Vérifiez la construction de l'arbre en valorisant des instruments simples et en les comparant à des valeurs analytiques.

### Valoriser l'option "chooser" dans l'arbre trinomial


## Montrer que l'option "chooser"peut être répliquée par un portefeuille statique, et calculer sa valeur analytiquement. Comparer vos deux estimations.

