---
title: "Programmation Stochastique avec Recours"
author: "P. Hénaff"
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  beamer_presentation:
    colortheme: dolphin
    theme: Montpellier
  slidy_presentation: default

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
  - \usepackage{subfig}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{xfrac}
  - \linespread{1.05}
  - \usepackage[utf8]{inputenc}
  - \usepackage{pgfplots}
  - \usepackage{tikz}
  - \usetikzlibrary{shapes}
  - \usetikzlibrary{external}
  - \usepgfplotslibrary{external}
  - \usepackage[ruled, vlined, linesnumbered]{algorithm2e}


bibliography: ../library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```




```{r, results='hide', message=FALSE}
  library(foreach)
  library(doParallel)
  library(kableExtra)
  library(linprog)
  library(gtools)
  library(pracma)
  library(kernlab)
  library(optiSolve)
  library(piqp)
  library(latex2exp)
```

## Arbre de décision

\scalebox{.7}{
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=8cm},
  level 2/.style={sibling distance=4cm},
  level 3/.style={sibling distance=2cm}]
  \node {1}
    child {node {2 high}
      child {node {4 high}
        child {node {S-1}}
        child {node {S-2}}
        }
      child {node {5 low}
        child {node {S-3}}
        child {node {S-4}}
      }
    }
    child {node {3 low}
      child {node {6 high}
        child {node {S-5}}
        child {node {S-6}}
        }
      child {node {7 low}
        child {node {S-7}}
        child {node {S-8}}
      }
    };
\end{tikzpicture}
}


## Solution


```{r, echo=FALSE}
Amat <- matrix(0, nrow=15, ncol=30)

sce_up = c(1.25, 1.14)
sce_down = c(1.06, 1.12)
sce = rbind(sce_up, sce_down)

Amat[1,1:2] = 1
Amat[2,1:2] = -sce_up; Amat[2,3:4] = 1
Amat[3,1:2] = -sce_down; Amat[3,5:6] = 1
Amat[4,3:4] = -sce_up; Amat[4,7:8] = 1
Amat[5,3:4] = -sce_down; Amat[5,9:10] = 1
Amat[6,5:6] = -sce_up; Amat[6,11:12] = 1
Amat[7,5:6] = -sce_down; Amat[7,13:14] = 1

Amat[8,7:8] = sce_up; Amat[8,15:16] = c(-1, 1)
Amat[9,7:8] = sce_down; Amat[9,17:18] = c(-1, 1)
Amat[10,9:10] = sce_up; Amat[10,19:20] = c(-1, 1)
Amat[11,9:10] = sce_down; Amat[11,21:22] = c(-1, 1)
Amat[12,11:12] = sce_up; Amat[12,23:24] = c(-1, 1)
Amat[13,11:12] = sce_down; Amat[13,25:26] = c(-1, 1)
Amat[14,13:14] = sce_up; Amat[14,27:28] = c(-1, 1)
Amat[15,13:14] = sce_down; Amat[15,29:30] = c(-1, 1)

bvec = matrix(0, nrow=15, ncol=1)

W.0 = 55000
G = 80000
r_short = 4/100
r_long = 1/100
prob = 1/8

bvec[1,1] = W.0
bvec[8:15,1] = G
cvec = matrix(0, nrow=1, ncol=30)
cvec[1, seq(15,30,by=2)] = r_long * prob
cvec[1, seq(16,30,by=2)] = -r_short * prob

bvec = as.vector(bvec)
cvec = as.vector(cvec)

res = solveLP(cvec, bvec, Amat, const.dir=rep("==", length(bvec)), lpSolve=TRUE, maximum=TRUE,verbose=1)

sol = matrix(0, nrow=7, ncol=3)
for(i in seq(7)) {
  sol[i,] = c(i, res$solution[(2*i-1):(2*i)])
}
colnames(sol) <- c("Node", "Stock", "Bond")
```

```{r, echo=FALSE}
kable(sol, caption="Optimal asset allocation at each node in the 3-period tree",
booktabs=T, digits=0) %>%
kable_styling(latex_options="hold_position")
```

The excess or shortfall at horizon $T$ for each scenario is displayed in Table \@ref(tab:es).

```{r es, echo=FALSE}
wealth <- matrix(0, nrow=8, ncol=3)
colnames(wealth) <- c("Scenario", "Shortfall", "Excess")
for(i in seq(8)) {
  wealth[i,] = c(i, res$solution[14+2*i], res$solution[14+2*i-1])
}
kable(wealth, caption="Shortfall or excess by scenario",
booktabs = T, digits=0)
```


## Solution Stochastique vs Miope

```{r, echo=FALSE}
return.index <- permutations(2,3,c(0,1), repeats.allowed = T)
return.high <- 1.25
return.low <- 1.06

idx <- rowSums(return.index)
return.by.scenario <- return.high ^ (3-idx) * return.low ^ idx
wealth.by.scenario <- W.0 * return.by.scenario
```

```{r ev, echo=FALSE}
options(knitr.kable.NA = "")
df = data.frame(Scenarios=seq(8),
W1.Short=-pmin(wealth.by.scenario-G, 0),
W1.Long = pmax(wealth.by.scenario-G, 0),
W2.Short=wealth[,2],
W2.Long=wealth[,3])
kable(replace(df, df==0, NA), col.names=c("Scenario", "Shortfall", "Surplus", "Shortfall", "Surplus"),
booktabs=T, digits=0,
caption="Expected Value vs. Recourse solutions of the asset allocation problem.") %>%
  kable_paper() %>%
  add_header_above(c(" "=1, "EV solution"=2, "Recourse solution"=2)
  )
```



```{r, echo=F}
expected.cost.ev = mean( -df$W1.Short*r_short + df$W1.Long*r_long)
expected.cost.sp = mean( -df$W2.Short*r_short + df$W2.Long*r_long)
VSS = expected.cost.sp - expected.cost.ev
```

## Lagrangien Augmenté

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{$\rho$, tol, $k_{\mbox{max}}$}
\Output{$x^*$}
\BlankLine
$k <- 0; \lambda_k <- 0$\;
\While{$k<k_{\mbox{max}}$ and $t > tol$}{
    $x_{k+1} <-  \underset{x}{\mathrm{argmin}} \phi_k(x)$\;
    $\lambda_{k+1} <- \lambda_k - \rho g(x_{k+1})$\;
    $k <- k+1$\;
    $t_1  <- ||\lambda_{k+1} - \lambda_k||$\;
    $t_2 <- ||g(x_{k+1}||$\;
    $t <- t_1 \mbox{and} t_2$\;
    }
\caption{Augmented Lagrangian algorithm}
\end{algorithm}


## Algo Progressive Hedging

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{$\rho$, tol, $i_{\mbox{max}}$}
\Output{$x^*$}
\BlankLine
$i <- 0; \lambda_i <- 0; \mbox{converged} <- False$\;
\While{$k<k_{\mbox{max}}$ and $! \mbox{converged}$}{
    Solve the $K$ subproblems \ref{eq:subproblems} to obtain $x^{i+1}_k, k = 1, \ldots, K$\;
    Compute  $\hat{x}^{i+1} = \sum_{k=1}^K \pi_k x^{i+1}_k$\;
    Update the multipliers: $\lambda^{i+1}_k = \lambda^i_k - \rho(x^{i+1}_k - \hat{x}^{i+1})$\;
    $i <- i+1$\;
    $t_1  <- ||\lambda^{i+1} - \lambda^i|| < \mbox{tol}_1$\;
    $t_2 <- ||g^1(x^{i+1}|| < \mbox{tol} < \mbox{tol}_2$\;
    $t_3 <- ||g^2(x^{i+1}|| < \mbox{tol} < \mbox{tol}_3$\;
    $\mbox{converged} <- t_1 \ \  \mbox{and} t_2 \ \ \mbox{and} t_3$\;
    }
\caption{Progressive Hedging algorithm}
\end{algorithm}


```{r, echo=FALSE}
P <- matrix(0, nrow=4, ncol=4)
P[3,3] = 1
P[4,4] = 1
c <- rep(0, 4)
G <- matrix(c(1, 1, 0, 0,
              -1, -3, -1, 0,
              -4, -2, 0, -1), nrow=3, byrow=TRUE)
h <- c(10, -25, -25)
x_lb <- rep(0, 4)
sol <- solve_piqp(P=P, c=c, G=G, h=h, x_lb=x_lb )
```

```{r, echo=F}
sub.prob <- function(x.hat, r, w, q) {
P <- zeros(3,3)
P[1,1] <- r
P[2,2] <- r
P[3,3] <- 2
c <- c(w - r*x.hat, 0)
G <- matrix(c(1, 1, 0,
              -q[1], -q[2], -1), nrow=2, byrow = TRUE)
h <- c(10, -25)
x_lb <- rep(0,3)
sol <- solve_piqp(P=P, c=c, G=G, h=h, x_lb = x_lb)
sol
}
```


```{r, echo=F}
q = matrix(c(1,3,4,2), nrow=2, byrow=TRUE)
w = matrix(0, nrow=2, ncol=2)
x.hat <- c(0,0)
x = matrix(0, nrow=2, ncol=2)
tol <- 1.0e-2
x[1,] <- sub.prob(x.hat, r=0, w[1,], q[1,])$x[1:2]
x[2,] <- sub.prob(x.hat, r=0, w[2,], q[2,])$x[1:2]
x.hat <- colMeans(x)

conver <- FALSE
iter.count <- 0
iter.max <- 100
r = 2
x.iter <- matrix(0, nrow=iter.max, ncol=2)

while(!conver & (iter.count < iter.max)) {
iter.count <- iter.count + 1
x.iter[iter.count,] = x.hat
x[1,] <- sub.prob(x.hat, r, w[1,], q[1,])$x[1:2]
x[2,] <- sub.prob(x.hat, r, w[2,], q[2,])$x[1:2]


x.hat.old <- x.hat
x.hat <- colMeans(x)
w.old <- w
w[1,] = w[1,] + r * (x[1,] - x.hat)
w[2,] = w[2,] + r * (x[2,] - x.hat)

# convergence test
test.1 <- norm(x.hat-x.hat.old, type="2")
test.2 <- norm(w - w.old, type="2")

conver <- (test.1 <= tol) & (test.2 <= tol)
}
```

## First Example

```{r, PHA-plot, echo=FALSE, out.width="70%",fig.cap="Progressive Hedging Iterations", fig.align='left'}
plot(seq(iter.count),x.iter[1:iter.count,1], ylim=c(2, 8), type="l", col="blue", xlab="Iterations",
     ylab=expression(X))
lines(seq(iter.count), x.iter[1:iter.count,2], type="l", col="red")
abline(h=2.5)
abline(h=7.5)
```


## Second example: A 3-stage optimization problem

```{r, echo=F}
nb.vars = 8
A <- matrix(0, nrow=4, ncol=nb.vars)
rownames(A) = c("t_0", "t_1", "t_2", "t_3")
colnames(A) = c("xA.1", "xB.1","xA.2", "xB.2","xA.3", "xB.3","M", "P")
W.0 = 55000
G = 80000
b = c(W.0, 0, 0, G)
r_short = 4
r_long = 1
tol <- 1

pha.sub <- function(sce, r, w, x.hat, r_short, r_long, b) {
  nb.vars=8
  nb.x = 6
A <- matrix(0, nrow=4, ncol=nb.vars)
A[1,1:2] = 1
A[2,1:2] = sce[1,]; A[2,3:4] = -1
A[3,3:4] = sce[2,]; A[3,5:6] = -1
A[4,5:6] = sce[3,]; A[4,7:8] = c(1, -1)
Q = pracma::zeros(nb.vars, nb.vars)
for(i in seq(nb.x)) {
  Q[i,i] = r
}
c = c(w - r * x.hat, r_short, -r_long)
x_lb = rep(0, nb.vars)
piqp::solve_piqp(P=Q, c=c, A=A, b=b, x_lb=x_lb)
}
```


```{r, echo=FALSE}
sce = rbind(sce_up, sce_up, sce_up)
nb.x = 6
nb.vars = 8
nb.scen = 8

x.hat = (55000/2)*rep(1, nb.x)
r = 0
w = rep(0, nb.x)
sol = pha.sub(sce, r, w, x.hat, r_short, r_long, b)
```
```{r, echo=FALSE}
scen = rbind(sce_up, sce_down)
m.scen = as.matrix(expand.grid(stage.2=c(1,2), stage.1=c(1,2), stage.0=c(1,2)))
m.scen = m.scen[,c(3,2,1)]

mean.x <- function(scen.list, i.var) {
  # Compute the mean value of variable i.var over the bundle of scenarios
  # scen.list
  tmp = vector()
  for(i in scen.list) {
    tmp = c(tmp, all_sol[[i]]$x[i.var])
  }
  mean(tmp)
}

calc.x.hat <- function() {
x.hat = matrix(0, nrow=8, ncol=6)
# stage 0
s = seq(8)
x.hat[s,1] = mean.x(s, 1)
x.hat[s,2] = mean.x(s, 2)
# stage 1
s = seq(4)
x.hat[s,3] = mean.x(s, 3)
x.hat[s,4] = mean.x(s, 4)
s = as.integer(c(5,6,7,8))
x.hat[s,3] = mean.x(s, 3)
x.hat[s,4] = mean.x(s, 4)
# stage 2
s = c(1,2)
x.hat[s,5] = mean.x(s, 5)
x.hat[s,6] = mean.x(s, 6)
s = c(3,4)
x.hat[s,5] = mean.x(s, 5)
x.hat[s,6] = mean.x(s, 6)
s = c(5,6)
x.hat[s,5] = mean.x(s, 5)
x.hat[s,6] = mean.x(s, 6)
s = c(7,8)
x.hat[s,5] = mean.x(s, 5)
x.hat[s,6] = mean.x(s, 6)
x.hat
}
```


```{r, echo=FALSE, cache=TRUE, warning=FALSE}
# Detect the number of available cores
nb.available.cores <- parallel::detectCores()
nb.cores <- 8
# print(paste("Using ", nb.cores, " cores out of ", nb.available.cores))

# Create a cluster and register
cluster <- parallel::makeCluster(nb.cores, type="PSOCK")
doParallel::registerDoParallel(cl=cluster)
```

```{r}
# Initialize with perfect foresight
w = matrix(0, nrow=8, ncol=6)
r = 0
x.hat = matrix(0, nrow=8, ncol=6)

run.subproblems <- function(foo, scen, m.scen, r,w,x.hat, r_short, r_long, b) {

all_sol <- foreach(i = seq(nrow(m.scen))) %dopar% {
  sce = scen[m.scen[i,],]
  foo(sce, r, w[i,], x.hat[i,], r_short, r_long, b)
}
all_sol
}

all_sol <- run.subproblems(pha.sub, scen, m.scen, r, w, x.hat,
r_short, r_long, b)
x.hat <- calc.x.hat()

conver <- FALSE
iter.count <- 0
iter.max <- 2000
r = .00002
x.iter = matrix(0, nrow=iter.max, ncol=4)
while(!conver & (iter.count < iter.max)) {
iter.count <- iter.count + 1
x.iter[iter.count,] = x.hat[1,1:4]

all_sol = run.subproblems(pha.sub, scen, m.scen, r, w, x.hat, r_short, r_long, b)

x.hat.old <- x.hat
x.hat <- calc.x.hat()
w.old <- w
for(i in seq(nrow(m.scen))) {
  w[i,] = w[i,] + r*(all_sol[[i]]$x[1:6]-x.hat[i,])
}

# convergence test
test.1 <- norm(x.hat-x.hat.old, type="2")
test.2 <- norm(w - w.old, type="2")

conver <- (test.1 <= tol) & (test.2 <= tol)
}
```


```{r, PHA-plot-1, echo=FALSE, out.width="70%", fig.cap="Progressive Hedging Iterations for the 3-stages asset allocation problem, showing the convergence of the first 4 variables.", fig.align='left'}
alpha <- 1:4
plot(seq(iter.count),x.iter[1:iter.count,1], ylim=c(0, 70000), type="l", col=alpha[1], xlab="Iterations",
     ylab=expression(X))
lines(seq(iter.count), x.iter[1:iter.count,2], type="l", col=alpha[2])
lines(seq(iter.count), x.iter[1:iter.count,3], type="l", col=alpha[3])
lines(seq(iter.count), x.iter[1:iter.count,4], type="l", col=alpha[4])
legend("topleft", legend=TeX(sprintf(r'($x_%d$)', alpha)),
lwd=1,
col=alpha)
```
