---
title: "Quantitative Finance"
subtitle: "Pricing under Historical Distributions"
author: "Patrick Hénaff"
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"

output:
  bookdown::pdf_document2:
    keep_tex: no
    fig_caption: yes
    latex_engine: pdflatex
csl: ../apa.csl
geometry: margin=1in
bibliography: ../library.bib
email: pa.henaff@gmail.com
fontfamily: mathpazo
fontsize: 11pt
header-includes:
  - \linespread{1.05}
  - \usepackage[utf8]{inputenc}
  - \usepackage{pgfplots}
  - \usepackage{tikz}
  - \usetikzlibrary{shapes}
  - \usetikzlibrary{external}
  - \usepgfplotslibrary{external}
  - \usepackage[ruled, vlined, linesnumbered]{algorithm2e}
  - \usepackage[utf8]{inputenc}
---

```{r folder-tools, include=TRUE, echo=FALSE}
get.src.folder <- function() {
  path.expand("../GP/src")
}
source(file.path(get.src.folder(), 'OptionUtils.R'))
```

In this note, we consider methods for pricing derivatives under a
historical density for the underlying asset.

In order to motivate this approach, consider 
Figure \@ref(fig:socal-basis) which represents the distribution of price for the “SoCal
basis”. This is the difference between the price of natural gas at Henry
Hub (the delivery point in Southern Louisiana for the NYMEX Futures
contract), and the price in Southern California. In general, the price
difference reflects the cost of transporting gas from East to West, say
less than a douzen cents per unit of volume. Once in a while, however, a
disruption in supply causes the price differential to jump to \$3 or \$5. In that context, what is the price of a call option struck at \$2?

```{r, socal-basis, echo=FALSE,fig.align='center', fig.cap="Distribution of So-Cal basis (USD per mmbtu)", out.width="50%"} 
knitr::include_graphics("../Graphics/lotteryTicket.pdf")
```

It is tempting to use make use of historical data, given the very specific nature of the distribution. The question then becomes:
Can we transform an empirical density into a density that:

1.  is risk-neutral,
2.  is consistent with current market observations, and
3.  retains the stylized facts of the historical density?

There is a abundant literature on the subject of derivative pricing
under empirical densities. See, for example, the review article of
Jackwerth [@Jackwerth99], as well as the paper by Derman and
Zou [@ZouJuly1999]. The principle of Derman’s approach is to start with
an empirical density for the underlying asset, and to adjust this
density in order to make it risk-neutral. Under the risk-neutral
density, the expected value of the settlement of a futures contract must
be the current forward price, and the fair value of a derivative is the
discounted expected value of the payoff.

How do we transform an empirical density into a risk-neutral one? The
intuition is to adjust the empirical density in a way
that “disturbs” it as little as possible. The concept of entropy is
useful to quantify the notion of closeness between distribution. 

# The Derman-Zou density adjustment

## Entropy

The entropy of a random variable $X$ is defined as

$$H(X) = -\sum_i p(x_i) \ln(p(x_i))$$

Therefore, the entropy of a uniform discrete variable with probability
$1/N$ associated with each value $x_i, i=1, \ldots, N$ is $\ln(N)$.

Let $P$ and $Q$ be two discreet random variables. The relative entropy of $P$ and
$Q$ measures the closeness between the probability distributions over the
same set:

\begin{align}
    S(P,Q) & =  E_Q\left\{\log Q - \log P\right\} \nonumber \\
    & =  \sum_x Q(x) \log \left(\frac{Q(x)}{P(x)}\right)
\end{align}

```{r, pq-dist, echo=FALSE, eval=TRUE, fig.margin=TRUE, fig.width=4,fig.height=4, fig.cap="Uniform distributions"}
  x <- seq(0,10,length=1000)
  u1 <- dunif(x, 0, 10)
  u2 <- dunif(x, 2, 8)
  u3 <- dunif(x, 4.5, 5.5)

  plot(x, u1, type='n', ylim=c(0,1), xlab="x", ylab="P(x)")
  polygon(c(10,0,x), c(0,0,u1), col='green')
  polygon(c(10,0,x), c(0,0,u2), col='red')
  polygon(c(10,0,x), c(0,0,u3), col='blue')

  text(1,.05, expression(P))
  text(3,.1,expression(Q[1]))
  text(5,.5, expression(Q[2]))
```

In order to form an intuition regarding relative entropy, consider a uniform distribution $P$ over
the set $X_N = (x_i, i=1, \ldots, N)$ and the distribution $Q$, also
uniform over a subset $X_M \in X_N$ of size $M$:

$$
q(x_i) =
    \left\{
        \begin{array}{rl}
             \frac{1}{M} & \mbox{if $x_i \in X_M$} \\
             0 & \mbox{otherwise}
        \end{array}
    \right.
$$

The distributions are represented in figure \@ref(fig:pq-dist). Then

$$S(Q,P) = \ln(N) - \ln(M)$$

Now write $M = N^\alpha$, the relative entropy can be expressed by:

$$S(Q,P) = \ln(N)(1 - \alpha)$$

The relative entropy is inversely proportional (in a log scale) to the
number of elements that are common to $X_M$ and $X_N$.


A natural procedure to compute the adjusted density $Q$ is to minimize the relative
entropy $S(P,Q)$ between the historical density $P$ and $Q$ under the constraint that $Q$ is risk-neutral. We can
expand upon this idea by computing $Q$ so that $Q$ satisfies an
arbitrary number of additional constraints, in particular that 
$Q$ prices exactly a number of benchmark derivatives.
Next section describes a method for performing this calculation.

## Computation of the risk-neutral density


Consider a density $P$ evaluated over $n$ intervals, so that:

$$p_i = P(x_i \leq X \leq x_{i+1})$$

We look for the discrete density $Q$ defined over the sample sampling,
which solves the following minimization problem:

$$
\begin{array}{ll}
\mbox{max} & -\sum_{i=1}^n p_i \log \frac{p_i}{q_i} \\
\mbox{such that}   \nonumber \\
& \sum_{i=1}^n p_i  =  1 \nonumber \\
& \sum_{i=1}^n p_i A_j(x_i)  =  b_j, j=i, \ldots, m  \nonumber
\end{array}
$$

Constraints of a second type may be used to calibrate the
risk-neutral density to match a variety of market information.

To match a given at-the-money volatility ($\sigma$), set:

\begin{align}
    A_j(x) & =  x^2 \nonumber \\
    b_j & =  \sigma^2 + \bar{x}^2
\end{align}

To match the price $s$ of a straddle, set:

\begin{align}
    A_j(x) &= \mid x-K \mid \nonumber \\
    b_j &= s e^{r(T-t)}
\end{align}

This formulation may be applied to an arbitrary set of options.

Let's write the Lagrangian and the first-order conditions:

$$
L = - \sum_i p_i \log \frac{p_i}{q_i} + \sum_j \lambda_j (y_j - \sum_i p_i
    A_j(x_i)) + \mu (1-\sum_i p_i)
$$


\begin{align}
  \frac{\partial L}{\partial p_i} &= -\ln(\frac{p_i}{q_i}) - 1 -\sum_j \lambda_j A_j(x_i) -\mu = 0 (\#eq:foc-1) \\
  \frac{\partial L}{\partial \lambda_j} &= y_j - \sum_i p_i A_j(x_i) = 0 (\#eq:foc-2) \\
  \frac{\partial L}{\partial \mu} &= 1-\sum_i p_i = 0 (\#eq:foc-3)
\end{align}

Using \@ref(eq:foc-1), we obtain,

$$p_i = q_i e^{(-1-\sum_j \lambda_j A_j(x_i) -\mu)}$$

Divide by $\sum_i p_i$ to eliminate $\mu$ and a constant term:

\begin{align}
    p_i &= \frac{q_i e^{(-\sum_j \lambda_j A_j(x_i) -\mu)}}{\sum_k q_k e^{(-1-\sum_j \lambda_j A_j(x_k) -\mu)}} \\
    &= \frac{q_i e^{(-\sum_j \lambda_j A_j(x_i))}}{\sum_k q_k e^{(-1-\sum_j \lambda_j A_j(x_k))}}
\end{align}

To calculate $\lambda_j$, we use \@ref(eq:foc-2):

$$
y_j =  \sum_i p_i A_j(x_i)
$$

and substitute the expression for $p_i$:

$$
y_j = \frac{\sum_i q_i e^{(-1-\sum_j \lambda_j A_j(x_i))} A_j(x_i)}{\sum_i q_i e^{(-1-\sum_j \lambda_j A_j(x_i))}}
$$

To summarize, the values for the adjusted probabilities $p_i$ are found to
be:

$$p_i = \frac{e^{-\sum_j \lambda_j A_j(x_k)}}{\sum_k e^{-\sum_j \lambda_j
    A_j(x_k)}}$$

with $\lambda_j$ solving the system:

$$
y_j = \frac{\sum_i q_i e^{-\sum_j \lambda_j A_j(x_i)} A_j(x_i)}{\sum_i q_i
e^{-\sum_j \lambda_j A_j(x_i)}}, j=1, \ldots, m
$$


The density can be further calibrated to available market information.
Figure \@ref(fig:atm-straddle) shows the result of the minimum entropy
adjustment to the sample distribution to match both the forward price
and the value of at-the-money straddles.

```{r atm-straddle, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Adjusted densities calibrated to straddles.', out.width='0.5\\linewidth'}
knitr::include_graphics("../Graphics/Straddle.pdf")
```

# Monte-Carlo pricing with historical paths

The principle here is to use actual historical paths in a Monte-Carlo
simulation. In this section, we summarize the model developed by
Potters, Bouchaud and Sestovic [@Marc2001].


Let’s first introduce some notation:

$x_k$
:   value of underlying asset at step $k$

$C_k(x_k)$
:   value of the derivative

$\phi_k(x_k)$
:   hedge ratio for derivative $C_k$ 

Define the local risk $R_k$ as:

$$E^P\left[(C_{k+1}(x_{k+1}) - C_k(x_k) - \phi_k(x_k)[x_{k+1} - x_k])^2\right]$$

where $E^P[]$ is the expectation under the objective probability
measure.

We look for the pricing function $C_k(x)$ that minimizes the residual
hedging risk.

The functions $C(x)$ and $\phi(x)$ are approximated by a set
of basis functions:

\begin{align}
    C_k(x) &= \sum_{a=1}^M \gamma_a^k C_a(x) \\
    \phi_k(x) &= \sum_{a=1}^M \gamma_a^k \frac{\partial C_a(x)}{\partial x}
\end{align}

Splines provide a convenient set of basis functions: given a set of
knots $t_i, i = 1 \ldots, k$, the polynomial spline of
degree $n$ is defined by:

$$
C(x) = \sum_{j=0}^n b_{0,j}x^j + \sum_{i=1}^k \sum_{j=0}^n
    b_{i,j}(x-t_i)_+^j
$$

Thus, a spline of degree $n$ with $k$ knots is a linear combination of
$m = (k+1)(n+1)$ basis functions. The derivative of $C(x)$ with respect to
$x$ is readily computed. To simplify notation, let:

\begin{align}
    C(x) &= \sum_{a=1}^m \beta_a F_a(x) \nonumber \\
    C'(x) &= \sum_{a=1}^m \beta_a F'_a(x) \nonumber\end{align}

where $F_a(x)$ are the elementary basis functions. At each step $t$ in the hedged
Monte-Carlo simulation, we obtain the price function by solving for
$\beta$ the following optimization problem (formulation for a call):

\begin{align}
\mbox{min} & \sum_{l=1}^N  [e^{-\rho} C_{t+1}(x^l_{t+1}) - \\
        & \sum_{a=1}^M \beta_a ( F_a(x_t^l) + F'_a(x_t^l) (x^l_{t+1} e^{-\rho} - x^l_t))]^2 \\
        \mbox{such that}  &  \notag \\
        & \sum_{a=1}^M \beta_a  F_a(x_t^l) >=  I(x_t^l) \notag \\
        & \sum_{a=1}^M \beta_a ( F_a(x_t^l) - F_a(x_t^{l-1})) sgn(x_t^l -
        x_t^{l-1}) >= 0 (\#eq:non-neg)
\end{align}

where $I(x)$ is the intrinsic value of the derivative being priced. The
second constraint enforces the basic non-arbitrage condition that the
value of a call must be a monotonically increasing function of the spot.

A simple modification of the model enables us to account for the bid-ask
spread on transactions. We assume that the price paths are mid-market
prices, and that transactions are negotiated at the bid or ask price.
Let $\epsilon$ be the bid-ask spread. The local risk becomes:

$$
(C_{k+1}(x_{k+1}) - C_k(x_k) - \phi_k(x_k) (e^{-\rho}
    x_{k+1} -x_k - \delta \epsilon/2))^2
$$

where

$$
\delta = \left\{ \begin{array}{ll} -1 & (x_k - e^{-\rho} x_{k+1})
    >=0 \\
    1 & (x_k - e^{-\rho} x_{k+1}) < 0 \end{array} \right.
$$


So far, we have only considered contracts with a single payoff at
expiry. A simple extension of the model can accommodate arbitrary
contingent cash flows at each step of the simulation. Assume that as
each time step, the contract generates a cash flow $F(x_k)$. We then
define the price function $C(x_k)$ as the contract value ex cash flow.
The local risk function is then:

$$
(C_{k+1}(x_{k+1}) + F_{k+1}(x_{k+1}) - C_k(x_k) + \phi_k(x_k) (x_k
    - e^{-\rho} x_{k+1} + \delta \epsilon/2))^2
$$

With contingent cash flows at each period, the constraints defined by
equation \@ref(eq:non-neg) need to be reformulated. Consider first the
constraint that the value of the contract must be greater or equal to
the intrinsic value of the European option. With multiple cash flows,
the equivalent constraint is that the contract value at a given time
step and state must be greater than the sum of the expected discounted
cash flows, under the conditional probability of this time step and
state. The rest of the algorithm is left unchanged.

# References {.unlisted .unnumbered}
