---
title: "Quantitative Finance"
subtitle: "Implied Tree"
author: "Patrick Hénaff"
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"

output:
  bookdown::pdf_document2:
    keep_tex: no
    fig_caption: yes
    latex_engine: pdflatex
csl: ../apa.csl
geometry: margin=1in
bibliography: ../library.bib
email: pa.henaff@gmail.com
fontfamily: mathpazo
fontsize: 11pt
header-includes:
- \linespread{1.05}
- \usepackage[utf8]{inputenc}
---

```{r folder-tools, include=TRUE}
get.src.folder <- function() {
  path.expand("../GP/src")
}
```

```{r setup-utils, include=TRUE}
source(file.path(get.src.folder(), 'OptionUtils.R'))
```

The Breeden-Litzenberger formula provides the density of the underlying
asset at expiry as a function of vanilla prices. In many instances, for
pricing American options for example, it is important to know the
density at any time $t$ between the current time and expiry. In this
note, we construct the function $\sigma(S_t,t)$, i.e. the
instantaneous volatility as a function of the underlying asset and time.

# The Derman-Kani Implied Tree

In a standard trinomial tree, the transition probabilities are
independent of the state. By contrast, the Derman-Kani implied tree
involves transition probabilities that are state-dependent. With this
extension, one obtain a model that matches the observed prices of
vanilla options at various maturities and strikes.

## Principle

We start with a trinomial tree constructed with a constant volatility.
This determines the geometry of the tree, according to the formulae
derived earlier. We will now modify the transition
probabilities in order to match observed option prices.

Let’s focus on one node and its three children nodes. The notation is
illustrated in Figure \@ref(fig:tree-diagram).

```{tikz, tree-diagram, fig.cap="One step of a trinomial tree", fig.align="center", echo=FALSE}
  \begin{tikzpicture}[level distance=1.5cm]
    \tikzstyle{level 0}=[rectangle, draw]
    \tikzstyle{level 1}=[rectangle, draw]
    \tikzstyle{edge from parent}=[->,draw]
    \tikzstyle{blank}=[rectangle]
    \node[level 0] {$S_i$} [grow'=right,sibling distance=1cm]
      child {node[level 1] {$S'_{i+2}$}
            edge from parent node[blank,above left] {$p_i$}
            }
      child {node[level 1] {$S'_{i+1}$}
            edge from parent node[blank,below left] {}
            }
      child {node[level 1] {$S'_i$}
            edge from parent node[blank,below left] {$q_i$}
            };
    \end{tikzpicture}
```

States $S_i$ are associated with maturity $T$, and states $S'_i$ with
maturity $T+\Delta t$. We will also use the following notation:

$\lambda_i$
:   state price for node $i$

$C(S'_i)$
:   call price, strike $S'_i$, maturity $T+\Delta t$

The call can be priced in the trinomial tree, giving:

$$\begin{aligned}
C(S'_i) &=& \lambda_i e^{-r\Delta t} p_i (S'_{i+2} - S'_{i+1}) \\
&& + \sum_{j=i+1}^{2n} \lambda_j e^{-r\Delta t} \left[ p_j (S'_{j+2} -S'_{i+1}) + (1-p_j-q_j)  (S'_{j+1} -S'_{i+1}) + q_j  (S'_{j} -S'_{i+1}) \right]\end{aligned}$$

The price process in the tree is a martingale, and thus:
$$\begin{aligned}
S'_j e^{r \Delta t} &=& F_j \\
&=& p_j S'_{j+2} + (1-p_j-q_j)  S'_{j+1} + q_j  S'_{j} -S'_{i+1}\end{aligned}$$

Using this identiy, the call price becomes:
$$\begin{aligned}
C(S'_i) &=& \lambda_i e^{-r\Delta t} p_i (S'_{i+2} - S'_{i+1}) \\
&& + \sum_{j=i+1}^{2n} \lambda_j e^{-r\Delta t} \left[ F_j - S'_{i+1} \right]\end{aligned}$$

In the equation above, all quantities except $p_i$ are known, yielding:
$$p_i = \frac{e^{r\Delta t} C(S'_{i+1}) - \sum_{j=i+1}^{2n} \lambda_j e^{-r\Delta t} \left[ F_j - S'_{i+1} \right]}{\lambda_i(S'_{i+2} - S'_{i+1})}
\label{eq:trinomial-pi}$$

Using again the martingale property, one can compute the probability
$q_i$:

$$q_i = \frac{F_i - p_i (S'_{i+2} - S'_{i+1}) - S'_{i+1}}{S'_i - S'_{i+1}}
\label{eq:trinomial-qi}$$

The corresponding local volatility at node $S'_i$ is finally obtained
by:
$$\begin{aligned}
\lefteqn{\sigma(S'_i, T) F_i^2 \Delta t =} \nonumber \\
&& p_i (S'_{i+2} - F_i)^2 + (1-p_i-q_i) (S'_{i+1} - F_i)^2 + q_i (S'_{i} - F_i)^2
\label{eq:local-vol}\end{aligned}$$

A similar calculation can be performed with put prices.

## Illustration

For the sake of this example, we will use a simple expression for the
Black-Scholes volatility, function of time to expiry ($T$) and spot
price ($S$):

$$\sigma(S, T) = \sigma_0 (1+a(S-100)) + bT
(\#eq:bs-vol-in-tree)$$

```{r, label=Black-Scholes-Vol-Function, echo=T}
  sigma <- .3
  a <- -.2/100
  b <- 0
  bsvol <- function(S,T) {
    sigma*(1+a*(S-100)) + b*T
  }
```

We now proceed step-by-step in the calculation process. The first step
is to construct a classical trinomial tree, with constant probabilities.
This determines the geometry of the tree.

In this example, we construct a 3-periods tree, anual volatility is
$30\%$. For simplicity, interest rate and dividend yield is set to 0.

```{r, label=ImpliedTree-2, echo=T}
  ## all rates = 0
  r <- 0
  b <- 0

  ## time steps
  n <- 3
  dt <- .01

  # number of rows and columns in the price grid
  nr <- 2*(n+1)-1
  nc <- (n+1)
  T <- seq(0, nc) * dt

  ## Trinomial tree
  u <- exp(sigma*sqrt(2*dt))
  d <- 1/u
  ## constant probabilities
  pu <- ((exp(r*dt/2)-exp(-sigma*sqrt(dt/2)))/(exp(sigma*sqrt(dt/2))-exp(-sigma*sqrt(dt/2))))^2
  pd <- (1-sqrt(pu))^2
  pm <- 1-pu-pd
  ## grid of prices at each tree node
  S <- matrix(data=NA, nrow=nr, ncol=nc)
  for(j in seq(1,nc)) {
    S[1:(1+2*(j-1)),j] <- 100 * u^(j-1) * d^seq(0, 2*(j-1))
  }
```

```{r, echo=FALSE}
TrinomialTreePlot <- function (TrinomialTreeValues, dx = -0.025, dy = 0.4, cex = 1,
    digits = 2,  ...)
{
    # draw 3 branches originating at node (x,y)
    drawLines <- function(x,y,col=2) {
      xx = c(x,x+1)
      for(k in -1:1) {
        yy = c(y, y+k)
       lines(x=xx,y=yy,col=col)
      }
    }
    Tree = round(TrinomialTreeValues, digits = digits)
    depth = ncol(Tree)
    # frame and coordinates:
    plot(x = c(1, depth), y = c(-depth+1, depth-1), type = "n",
        col = 0, yaxt='n', xaxt='n', xlab='step', ylab='', ...)
    axis(1,at=1:depth)
    # tree root
    points(x = 1, y = 0)
    drawLines(1,0)
    text(1 + dx, 0 + dy, deparse(Tree[1, 1]), cex = cex)
    for (i in 1:(depth - 1)) {
        y = seq(from = i, by = -1, length = 2*i + 1)
        x = rep(i, times = length(y)) + 1
        points(x, y, col = 1)
        # place text
        for (j in 1:length(x)) text(x[j] + dx, y[j] + dy, deparse(Tree[j, i + 1]), cex = cex)

        if(i<(depth-1)) {
          for (k in 1:length(x)) drawLines(x[k], y[k])
        }
    }
    invisible()
}
```

The resulting tree is pictured in Figure \@ref(fig:Derman-Kani-1). This is
a standard trinomial tree. Branching probabilities are identical at each
node, volatility at each node is also constant, $30\%$ annual rate.

```{r Derman-Kani-1,echo=FALSE,fig.width=11,fig.height=7, fig.cap="Constant volatility trinomial tree"}
  TrinomialTreePlot(S, dx = -0.025, dy = 0.4, cex = 1,
      digits = 1, main='Constant Volatility Tree')
```

The next step is to price calls and puts expiring at each time step, and
struck at each mode value. The volatility used to price each option is
time and strike dependent, per equation (\@ref(eq:bs-vol-in-tree).

The volatility grid is computed by evaluating the volatility function
for each node:

```{r, echo=TRUE}
  Vol <- matrix(data=NA, nrow=nr, ncol=nc)
  for(i in seq(1,nr)) {
    for(j in seq(1, nc))
      if(!is.na(S[i,j])) Vol[i,j] <- bsvol(S[i,j], T[j])
  }
```

and this volatility is used to compute the call and put prices expiring
at each time step and strike level:

```{r, }
  ## matrice des prix call/put
  Call <- matrix(data=NA, nrow=nr, ncol=nc)
  Put <- matrix(data=NA, nrow=nr, ncol=nc)
  for(i in seq(1,nr)) {
    for(j in seq(2, nc))
      if(!is.na(S[i,j])) {
          Call[i,j] <- CRRTrinomial('ce', S[1,1], S[i,j], T[j], r, b, Vol[i,j], j-1)$price
          Put[i,j] <- CRRTrinomial('pe', S[1,1], S[i,j], T[j], r, b, Vol[i,j], j-1)$price
      }
  }
```

Next, we use equations \@ref(eq:trinomial-pi) and \@ref(eq:trinomial-qi)
to compute the transition probabilities at each node. The probability
$p_i$ is computed by:

```{r, }
  p <- function(i, j) {
    # S'_{i+1} and S'_{i+2}
    SP1 <- S[i+1, j+1]
    SP2 <- S[i, j+1]
    # vector of lambdas
    tmp = 0
    if(i>1) {
      l <- Lambda[1:(i-1),j]
      F <- S[1:(i-1),j]*exp(r*dt)
      #print(paste('F: ', F, ' SP1: ', SP1, ' SP2: ', SP2))
      tmp = t(l) %*% (F-SP1)
    }
    # prob
    (exp(r*dt)*Call[i+1,j+1] - tmp)/(Lambda[i,j]*(SP2-SP1))
  }
```

and the probability $q_i$ is determined by:

```{r, }
  q <- function(i, j) {
    # S'_{i+2}, S'_{i+1} and S'_{i}
    SP2 <- S[i, j+1]
    SP1 <- S[i+1, j+1]
    SP <- S[i+2, j+1]
    F <- S[i,j]*exp(r*dt)
    (F-p(i,j)*(SP2-SP1)-SP1)/(SP-SP1)
  }
```

With these two functions, we can proceed recursively, computing the
state prices and the transition probabilities one time step at a time.
Note that functions $p$ and $q$ above use as input the state prices up
to time step $i-1$ in order to compute the transition probabilities at
time step $i$. The transition probabilites for the root node are
computed immediately:

```{r, }
  Lambda <- matrix(data=NA, nrow=7, ncol=4)
  Lambda[1,1] <- 1
  Pu <- p(1,1)
  Pd <- q(1,1)
  Pm <- 1-Pu-Pd
```

and this provides the data for computing the state prices
$\lambda_{i,2}, i=1, \ldots, 3$ for time step $\Delta t$.

```{r, }
  Lambda[1,2] <- Pu * exp(-r*dt)
  Lambda[2,2] <- Pm * exp(-r*dt)
  Lambda[3,2] <- Pd * exp(-r*dt)
```

The state prices for the other time steps are computed similarly.

```{r, }
  Lambda[1,3] <- p(1,2)*Lambda[1,2]
  Lambda[2,3] <- (1-p(1,2)-q(1,2))*Lambda[1,2] + p(2,2)*Lambda[2,2]
  Lambda[3,3] <- q(1,2)*Lambda[1,2] + (1-p(2,2)-q(2,2))*Lambda[2,2] + p(3,2)*Lambda[3,2]
  Lambda[4,3] <- (1-p(3,2)-q(3,2))*Lambda[3,2] + q(2,2)*Lambda[2,2]
  Lambda[5,3] <- q(3,2)*Lambda[3,2]

  Lambda[1,4] <- p(1,3)*Lambda[1,3]
  Lambda[2,4] <- (1-p(1,3)-q(1,3))*Lambda[1,3] + p(2,3)*Lambda[2,3]
  Lambda[3,4] <- q(1,3)*Lambda[1,3] + (1-p(2,3)-q(2,3))*Lambda[2,3] + p(3,3)*Lambda[3,3]
  Lambda[4,4] <- q(2,3)*Lambda[2,3] + (1-p(3,3)-q(3,3))*Lambda[3,3] + p(4,3)*Lambda[4,3]
  Lambda[5,4] <- q(3,3)*Lambda[3,3] + (1-p(4,3)-q(4,3))*Lambda[4,3] + p(5,3)*Lambda[5,3]
  Lambda[6,4] <- (1-p(5,3)-q(5,3))*Lambda[5,3] + q(4,3)*Lambda[4,3]
  Lambda[7,4] <- q(5,3)*Lambda[5,3]
```

Since interest rate is 0, the state prices at each time step should sum
up to 1. This is verified by:

```{r, }
  z <- apply(Lambda, 2, function(x){sum(x[!is.na(x)])})
  print(z)
```

Having determined the state prices, we record the transition
probabilites in grids, in order to facilitate the display. The up and
down probabilities associated with each node are displayed in
Figure \@ref(fig:Derman-Kani-3).

```{r, }
  Pup <- matrix(data=NA, nrow=nr, ncol=nc)
  Pdn <- matrix(data=NA, nrow=nr, ncol=nc)
  Pmd <- matrix(data=NA, nrow=nr, ncol=nc)
  for(i in seq(1,nr)) {
    for(j in seq(1, nc-1))
      if(!is.na(S[i,j])) {
        Pup[i,j] <- p(i,j)
        Pdn[i,j] <- q(i,j)
        Pmd[i,j] <- 1-Pup[i,j]-Pdn[i,j]
      }
  }
```

```{r Derman-Kani-3,echo=FALSE,fig.width=11,fig.height=7, fig.cap="Up and down probabilities"}
  par(mfrow=c(1,2))
  TrinomialTreePlot(Pup[,1:(nc-1)], dx = -0.025, dy = 0.4, cex = 1,
      digits = 2, main='Prob Up', ylim=c(-(nc-2), nc-1))
  TrinomialTreePlot(Pdn[,1:(nc-1)], dx = -0.025, dy = 0.4, cex = 1,
      digits = 2, main='Prob down', ylim=c(-(nc-2), nc-1))
  par(mfrow=c(1,1))
```

Finally, the local volatility at each node can be computed from the
transition probabilities:

```{r, }
  lvol <- function(i,j) {
    SP2 <- S[i, j+1]
    SP1 <- S[i+1, j+1]
    SP <- S[i+2, j+1]
    F <- S[i,j]*exp(r*dt)
    sqrt((Pup[i,j]*(SP2-F)^2 + Pdn[i,j]*(SP-F)^2 + Pmd[i,j]*(SP1-F)^2)/(F^2*dt))
  }

  LVol <- matrix(data=NA, nrow=nr, ncol=nc)
  for(i in seq(1,nr)) {
    for(j in seq(1, nc-1))
      if(!is.na(S[i,j])) {
        LVol[i,j] <- lvol(i,j)
      }
  }
```

Figure \@ref(fig:Derman-Kani-4) shows the local volatility associated with
each node, and, for comparison, the Black-Scholes volatility, which is
the average volatility from time $0$ to expiry associated with a
particular strike and expiry date.

```{r Derman-Kani-4,echo=FALSE,fig.width=11,fig.height=7, fig.cap="Local volatility and Black-Scholes average volatility at each node"}
  par(mfrow=c(1,2))
  TrinomialTreePlot(LVol[,1:(nc-1)]*100, dx = -0.025, dy = 0.4, cex = 1,
      digits = 1, main='Local Vol', ylim=c(-(nc-2), nc-1))
  TrinomialTreePlot(Vol[,1:(nc)]*100, dx = -0.025, dy = 0.4, cex = 1,
      digits = 1, main='BS Vol', ylim=c(-(nc-1), nc))
  par(mfrow=c(1,1))
```
