---
title: "Finance Quantitative"
subtitle: "Risque de Taux"
author: Patrick Hénaff
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    keep_tex: false
    fig_caption: yes
    latex_engine: pdflatex
geometry: margin=1in

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{eurosym}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include=FALSE, echo=TRUE}
library(quantmod)
library(xts)
library(hornpa)
library(lubridate)
library(xtable)
library(PerformanceAnalytics)
library(TTR)
library(SuppDists)
library(lubridate)
library(roll)
library(Hmisc)
library(nFactors)
library(kableExtra)
library(quadprog)
```


# Relative Price Change (MPP)

Calculate the percentage price change for 4 bonds with different annual coupon
rates (5\% and 10\%) and different maturities (3 years and 10 years), 
starting with
a common 7.5\% YTM (with annual compounding frequency), 
and assuming successively a new yield of 5\%, 7\%, 7.49\%, 7.51\%, 8\% and 10\%.

# Treasury Bond

Suppose that the Treasury department issues a new 2-year bond that settles today
and matures in exactly 2 years. It has a yield of 6\% and a coupon rate of 6\%.
Coupon frequency and compounding frequency are assumed to be semiannual.
There are 182 days in the first coupon period. Answer the following questions.

\begin{enumerate}
\item What is the price of the bond?
\item What is the accrued interest?
\item What is the duration and modified duration?
\end{enumerate}  

# Duration

\begin{enumerate}
  \item Find the duration of a 10 year, $8\%$ bond trading at par.
\item Analytical formula for duration.
\begin{enumerate}
\item Using the formula for bond price:
  \[
  P = \frac{c}{r}[1-(1+r)^{-n}] + (1+r)^{-n}
  \]
  compute $\frac{\partial \ln(P)}{\partial r}$
  \item Use this result to derive a formula for duration
  \item Show that the limiting value of duration as maturity is increased to infinity is:
\[
\lim_{n \rightarrow \infty} D = \frac{1+r}{r}
\]
\end{enumerate}
\end{enumerate}

# Cash flow matching
  
Suppose the price is \$212 for a 2-year coupon bond with face of \$200 and an
annual coupon (first one is one year from now) of \$40. 
Suppose also that the price
is \$150 for a 1-year coupon bond with face of \$150 and an annual coupon (one
remaining, one year from now) of \$15.

Remaining pension benefits in a plan having two more years to go are \$95,000
one year from now and \$60,000 two years from now. 

\begin{enumerate}
\item What replicating portfolio of the two coupon bonds covers the pension liabilities exactly? 
  \item What is the price of the replicating portfolio?
  \end{enumerate}
  
# Bond dedication
  
In this problem, we construct a bond portfolio that generates a cash flow stream that matches a liability.

Assume that you must pay the amounts summarized in table~\ref{tab:liability}. You can invest in a portfolio of 5 bonds described in table~\ref{tab:bonds-2}.

  \begin{table}[h]
    \centering
    \begin{tabular}{ll}
    Year & Cash Flow \\
    \hline
    1    & -100 \\
    2    & -200 \\
    3    & -150 \\
    4 & -400 \\
    5 & -300
  \end{tabular}
    \caption{Liability cash flow stream}
    \label{tab:liability}
  \end{table}


  \begin{table}[h]
    \centering
    \begin{tabular}{llll}
    Bond & Maturity & Coupon & Yield \\
    \hline
    A & 1 & .05 & .05 \\
    B & 2    & .07 & .075 \\
    C & 3    & .06 & .058 \\
    D & 4 &  .05  & .049 \\
    E & 5 & .08 & .081
  \end{tabular}
    \caption{Available bonds for dedication}
    \label{tab:bonds-2}
  \end{table}

At every period, we can re-invest excess cash flow at a rate of $.02$, but cannot borrow.

Let's define the following notation:

\begin{description}
\item[$q_i$] quantity of bond $i$
\item[$C(t)$] cash balance at time $t$
  \item[$F_i(t)$] cash flow from 1 unit of bond $i$ at time $t$.    
\end{description}

The purpose of the problem is to determine $q_i, i \in {A, B, C, D}$ and $C(t), t=0, \ldots, 4$. The cash balance at end of year 5, $C(5)$ should be 0.

\begin{enumerate}
  \item Write the accounting identity defining the cash-flow balance at each period (i.e. the balance between the money received and the money paid out). 
    \item Write the constrains on the variables $q_i$ and $C(t)$.
\item Your goal is to minimize the cost of this dedication strategy. Write the corresponding objective function.
\item Use the linprog package in R to solve the problem
\end{enumerate}


# Reinvestment Risk and Market Risk
  
Consider a 3-year standard bond with a 6\% YTM and a 100 \euro\ face value, which
delivers a 10\% coupon rate. Coupon frequency and compounding frequency are
assumed to be annual. Its price is 110.69 \euro\ and its duration is equal to 2.75. We assume that YTM changes instantaneously to become 5\%, 5.5\%, 6.5\% or 7\% and
stays at this level during the life of the bond. Whatever the change in this YTM,
show that the sum of the bond price and the reinvested coupons after 2.75 years
is always the same.

# Duration Hedging

An investor holds 100,000 units of bond A whose features are summarized in the
following table. He wishes to be hedged against a rise in interest rates by selling some bond H.

\begin{table}[ht]
  \centering
  \begin{tabular}{llllll}
\toprule \\
Bond & Maturity & Coupon rate (\%) & YTM (\%) & Duration & Price \\
\midrule \\
A & 18 & 9.5 & 8 &  9.5055 & 114.181 \\
H & 20 & 10 & 8 & 9.87 & 119.792 \\
\bottomrule
\end{tabular}
\end{table}

Coupon frequency and compounding frequency are assumed to be semiannual.
YTM stands for yield to maturity. The YTM curve is flat at an 8\% level.

\begin{enumerate}
  \item What is the quantity of the hedging instrument H that the investor has to sell?
\item We suppose that the YTM curve increases instantaneously by 0.1\%.
\begin{enumerate}
  \item What happens if the bond portfolio has not been hedged?
  \item And if it has been hedged?
  \end{enumerate}
  \item Same question as the previous one when the YTM curve increases instantaneously by 2\%.
\item Conclude.
\end{enumerate}

What is the quantity of the hedging instrument H that the investor has to sell?

# Rich-Cheap Bond Strategy
  
A trader implements a duration-neutral strategy that consists in buying a cheap
bond and selling a rich bond. Today, the
rich and cheap bonds have the following characteristics:

\begin{table}[ht]
  \centering
  \begin{tabular}{llll}
\toprule \\
Bond & Maturity & Coupon rate (\%) & YTM (\%) \\
\midrule \\
Rich & 10 & 5 & 7.50 \\
Cheap & 12 & 5.5 & 7.55 \\
\bottomrule
\end{tabular}
\end{table}

Coupon frequency and compounding frequency are assumed to be annual. Face
value is 100 \euro\  for the two bonds.

Compute the PV01 of the two bonds and find the hedged position.

  
