---
title: "Finance Quantitative"
subtitle: "Modèle Black-Scholes"
author: Patrick Hénaff
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    keep_tex: false
    fig_caption: yes
    latex_engine: pdflatex
geometry: margin=1in

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{eurosym}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include=FALSE, echo=TRUE}
library(quantmod)
library(xts)
library(hornpa)
library(lubridate)
library(xtable)
library(PerformanceAnalytics)
library(TTR)
library(SuppDists)
library(lubridate)
library(roll)
library(Hmisc)
library(nFactors)
library(kableExtra)
library(quadprog)
```

# Black-Scholes Model (1)
  

\begin{table}[ht]
  \centering
  \begin{tabular}{cc}
    \toprule
    Type & Call \\
    Strike & 50 \\
    Spot & 55 \\
    Maturity & 3 months \\
    Interest rate & 3\% \\
    Dividend yield & 0\% \\
    Volatility & .30 \\
    \bottomrule
  \end{tabular}
  \caption{Characteristics of an European option}
  \label{tab:opt}
\end{table}

- Use the Black-Scholes model to price the option described in Table \ref{tab:opt}.
- Assumes that a trader sells this option, and intends to delta-hedge her position. Compute the transactions that she will initiate to create her hedge portfolio.

# Black-Scholes Model (2)

On March 21, 2012, GOOG quotes \$636.91. Table \ref{tab:goog} provides the prices of selected options expiring on 18 Jan 2013:

\begin{table}[ht]
  \centering
  \begin{tabular}{ccc}
    \toprule
    Type & Strike & Price \\
    \midrule
    Call  & 635 & 60.70 \\
    Put   & 635 & 59.70 \\
    \bottomrule
  \end{tabular}
  \caption{Prices of options on GOOG}
  \label{tab:goog}
\end{table}

Assume an interest rate of $1.0\%$. Google does not pay any dividend. 

\begin{itemize}
  \item Compute the implied volatility for the call and the put. Comment your results.
    \item Use this result to price a call with strike 650.
    \end{itemize}
    
